---
title: "PUBP 614: Assignment 7"
author: "Elizabeth Mohney"
output: html_document
---

# Instructions
The goal of this assignment is to improve one of your earlier plots. You should improve the plot in
2 ways:
  • Change the visualization type
  • Use theme_ and/or scale_ to improve its appearance
  • Add tooltips (optional but recommended for practice)
The idea is that this assignment will help you produce a visualization that is very close to a final
product, one that is refined in its appearance and the amount and type of information it conveys.
You should make at least 3 theme_ and/or scale_ changes. These can involve the appearance of the
plot area, the axes, the font, etc., etc., etc.

# Code 
```{r warning=FALSE, include=FALSE}
library(here)
library(tidyverse)
library(maps)
library(ggthemes)     # For theme_map()
library(ggimage)
library(plotly)
library(htmlwidgets)
library(readxl)
```


```{r}
mortality <- read_excel(here("Datasets", "global_mortality.xlsx"))

mortality$country[mortality$country == "United States"] = "USA"
mortality$country[mortality$country == "United Kingdom"] = "UK"
mortality$country[mortality$country == "Cote d'Ivoire"] = "Ivory Coast"
mortality$country[mortality$country == "Democratic Republic of Congo"] = "Democratic Republic of the Congo"
mortality$country[mortality$country == "Congo"] = "Republic of Congo"
mortality$country[mortality$country == "Macedonia"] = "North Macedonia" #match names to datasets


world_map <- map_data("world") 
```


```{r}
mortality$region <- mortality$country # Merge the datasets
world_mort <- left_join(world_map, # dataset 1
                  mortality, # dataset 2
                  by = "region") # key variable(s) which can be more then one
```

```{r}
#Help Figure out what countries aren't matched
#world_mort_na <-  world_mort %>%
    #filter(is.na(`Cardiovascular diseases (%)`))
#table(world_mort_na$region)
```

```{r}
world_mort_up <-  world_mort %>%
    filter(year==2016 | year==1990)
```

```{r}

breaks <- c(0, 10, 20, 30, 40, 50, 60, 70) #add bin numbers

p <- ggplot(world_mort_up, aes(x = long, y = lat, group = group,
                                fill = cut(`Cardiovascular diseases (%)`, breaks = breaks)
                               )) +
  geom_polygon(color = "black", linewidth = .1, aes(text= paste(" Country Name:", country,"\n Percentage of Deaths:", `Cardiovascular diseases (%)`, "%" ))) +
  scale_fill_manual(values = c("#FFFFFF", "#FFEBEB", "#FFD5D5", "#FFB9B9", "#FF9D9D", "#FF8181", "#FF0000"),#add colors for bins 
                    breaks = levels(cut(world_mort_up$`Cardiovascular diseases (%)`, breaks = breaks)), 
                    labels = c("0-10", "10-20", "20-30", "30-40", "40-50", "50-60", "60-70")) +
  labs(title = "World Map of the Percentage of Deaths from Cardiovascular Disease By Country in 1990 and 2016",
       fill = "Percentage Rate") +
  facet_wrap(~year, ncol = 1) +
  ggthemes::theme_map() +
  theme(panel.background = element_rect(fill = "#0077be"), #turn sea to be blue to better understnad color 
        text = element_text(family = "Times New Roman", vjust = 0.5))

ggplotly(p, tooltips=text)
final <-ggplotly(p, tooltips=text)

#saveWidget(final, file = "assignment_7_new.html")
```

# Reflections
  I decided to switch from an animated plot to just a facet wrap of the first and last year recorded after the class critique. The other major critique that I received was the differences in region harder to see so I changed the continuous color gradient to a binned option increasing by 10. From there I added tool tips to be able to tell the country names and exact percentages. However, I was unable to get rid of the first line of the tool tips even when manually suppressing all tool tips. I then made some aesthetic changes by adding a blue background to represent the sea and changing the font to times new roman. 